---
title: "R_831_gettingAcquaintedWithDataset"
output: html_document
---

The package 'alr4' is needed:
```{r alr4, echo=TRUE,message=FALSE}
if(!require(alr4)) install.packages("alr4",repos = "http://cran.us.r-project.org")
library(alr4)
```

Getting acquainted with a standard dataset: UN11
```{r,echo=TRUE}
head(UN11,10)
summary(UN11)

```


Notice that per person GDP ranges over several orders of magnitude, so a log transformation is probably appropriate.  What kinds of relationships do we see among the variables on a quick inspection?

```{r, echo=TRUE}
pairs(~fertility+log(ppgdp)+pctUrban+lifeExpF,UN11)
```
Strong positive relationships between urbanization, ppgdp, and life expectancy, with fertility having a negative relationship to the others.  

Which country has the highest ppgdp?
```{r, echo=TRUE}
stripchart(UN11$ppgdp)
head(UN11[order(-UN11$ppgdp),],10)
UN11[which.min(UN11$ppgdp),]
```



Basic plotting (in this case plotting against log(ppgdp) is appropriate):
```{r, echo=TRUE}
plot(fertility~ppgdp,data=UN11)
plot(fertility~log(ppgdp),data=UN11)
UN11$logppgdp<-log(UN11$ppgdp)
```

Using lowess smoothing to aid visualization:
```{r,echo=TRUE}
plot(fertility~logppgdp,UN11)
# lowess(UN11$fertility~log(UN11$ppgdp))
lines(lowess(UN11$fertility~log(UN11$ppgdp)))
```


